
```{r include=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE)
```

## 2.5 Multiple Regression {#c2s5 .first}

Let us try a multiple regression model with linear and additive effects
of social setting and program effort:

```{r}
    fpe <- read.table("https://grodri.github.io/datasets/effort.dat")
m1 <- lm(change ~ setting, data=fpe) # redone for comparison
m2 <- lm(change ~ setting + effort, data=fpe)
summary(m2)
anova(m2)
```

This time we went directly to the `summary()` and `anova()`.
Compare the coefficients with Table 2.5 and the anova with Table 2.7.

The estimates suggest that fertility decline increases with social setting,
about a quarter of a percentage point per point of setting, when we compare
countries with the same effort. Similarly, fertility declines about one
percentage point more per point of program effort, when we compare countries
with the same social setting.

It is important to keep in mind that both conclusions are based on the model,
which assumes linearity and additivity. In fact, we can't really compare
countries which differ exactly by one point in effort and have the same setting,
so goodness of fit will be a central concern. You may also note that I try to
avoid using causal language when dealing with observational data.

The F-test for a linear effect of effort after taking into account setting is
18.5 on 1 and 17 d.f., and is the same as the square of the t-statistic of 4.3
on 17 d.f.

### Partial Correlations

Make sure you know how to reproduce the hierarchical anova in Table 2.8
in the notes. Let us check the partial correlation listed on page 27.
Here's the coefficient for change and effort controlling for setting:

```{r}
rss <- function(m) sum(residuals(m)^2)
sqrt(1 - rss(m2)/rss(m1))
```

Another way to obtain this coefficient is as the correlation between the
residuals of change given setting and effort given setting:

```{r}
cor(resid(lm(change~setting, data=fpe)), 
        resid(lm(effort~setting, data=fpe)))
```

Squaring this coefficient we find that program effort is associated with
about half the variation left in fertility decline in countries with the
same social setting. (It is common to say "after controlling for social
setting", but of course we are not able to manipulate any of these variables.)

<small>Updated fall 2022</small>
