
```{r include=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE)
```

## 2.6 One-Way Analysis of Variance {#c2s6 .first}

Let us read the data again, and then group social setting into three categories:
< 70, 70-79 and 80+.

We will use the `dplyr` package for data manipulation, so make sure you have
installed it. 
To recode setting we will use `mutate()` with the function `cut()`, which 
takes as  arguments the original variable and a set of cutpoints or "breaks", here 
the minimum (23), 70, 80, and the maximum (91). By default, the cutpoint itself is 
included in the lower category, so the intervals are closed on the right, as in 
(70,80], but we want the opposite, as in [70,80), so we specify `right=FALSE`. 
To make sure we include the highest cutpoint in the last category we specify 
`include.lowest=TRUE`. (I know, this option can be confusing; by default it 
will include the lowest cutpoint in the first category, as the name suggests, 
but having specified that `right` is false, it will actually include the 
highest cutpoint in the last category.)

R generates labels for the categories, and you should probably accept them at 
first to ensure that the grouping was done as intended, but I'll call the 
categories "low", medium", and "high", for consistency with the notes. We also
add this variable to the data frame.

```{r}
library(dplyr)
fpe <- read.table("https://grodri.github.io/datasets/effort.dat")
fpe <- mutate(fpe, setting.g = cut(setting, 
  breaks=c(min(setting), 70, 80, max(setting)), 
  right=FALSE, include.lowest=TRUE, labels=c("Low", "Medium", "High")))
```

To verify the grouping we need to split the data by categories of social setting 
and compute the range of setting in each group. 

```{r}
group_by(fpe, setting.g) |> summarize(bot=min(setting), top=max(setting))
```

Here `group_by()` splits `fpe` by categories of `setting.g`. The pipe operator `|>`
passes the result to `summarize()`, which computes the min and max of `setting`.
We see that the ranges included in each category are 35-68, 70-77 and 83-91, so 
obviously the cut was correct. We can use the same approach to compute the average 
fertility decline in each category of social setting:

Let us look at the mean response by level of social setting

```{r}
group_by(fpe, setting.g) |> summarize(change = mean(change))
```

We observe substantially more fertility decline in countries with higher
setting, but only a small difference between the low and medium
categories.

### A One-Factor Model

To fit a linear model treating social setting as a factor we simply use the 
categorical variable in the model formula. R will automatically create dummy 
variables for each category other than the first.

```{r}
m1g <- lm(change ~ setting.g, data=fpe)
summary(m1g)
anova(m1g)
```

Fertility declined, on average, 16 percentage points more in countries
with high setting than in countries with low setting. Compare the parameter
estimates with the values in Table 2.11 and the anova with Table 2.12 in the notes.

You can verify that the constant is the average decline in low setting countries,
and the coefficients for medium and high are the differences between medium and
low, and between high and low; in other words, differences with respect to the
omitted category.

We could, of course, compute the dummy variables "by hand", with the same results. 
Below I use `as.numeric()` to coerce a logical expression to take the values 1 and 0, 
otherwise the coefficients would have names like `settingMediumTRUE`.

```
fpe <- mutate(fpe, settingMedium = as.numeric(setting.g=="Medium"), 
  settingHigh=as.numeric(setting.g=="High"))
lm(change ~ settingMedium + settingHigh, data=fpe)
```

Using the factor notation is not only simpler but tells R that the terms belong together.

### The F-Test

The t-statistics produced by `summary()` compare each category to the reference 
cell. To obtain an overall test of the significance of social setting we need to 
compare the models with and without setting. This can be done using the `anova()` 
function:

```{r}
anova(m1g)
```

The F-test of 6.97 on 2 and 17 d.f. tells us that the differences between the
social setting categories are much larger than one would expect by chance if all
experienced the same decline in fertility.

It may be instructive to calculate the test "by hand" using the residual sum of 
squares for the null model and the model treating setting as a factor:

```{r}
rss <- function(lm) sum(residuals(lm)^2)
m0 <- lm(change ~ 1, data=fpe)
((rss(m0)-rss(m1g))/2)/(rss(m1g)/df.residual(m1g))
```

We obtain the same 6.97 on 2 and 17 d.f.
We can also compute the test from the coefficients as their variance-covariance 
matrix, as shown on page 32 of the notes. We do this using `coef()` to extract 
the coefficients and `vcov()` for their variance-covariance matrix, using -1 to 
exclude the constant:

```{r}
b = coef(m1g)[-1]
V = vcov(m1g)[-1,-1]
W = t(b) %*% solve(V) %*% b
c(W, W/2)
```

We obtain a Wald statistic of 13.94 on 2 d.f. in agreement with the notes. Dividing 
by 2 we obtain, once again, an F statistic of 6.92 on 2 and 17 d.f. Recall that 
the Wald statistic is asymptotically chi-squared, whereas under normality the F 
statistic has an F distribution.

*Exercise*: Obtain the parameter estimates and anova table for the model
with family planning effort grouped into three categories: 0-4, 5-14 and
15+, labelled weak, moderate and strong.

<small>Updated fall 2022</small>
