
```{r include=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE)
```

# 6. Multinomial Response Models {#c6s1 .first}

This section deals with regression models for discrete data with more
than two response categories, where the assumption of a multinomial
distribution is appropriate. We will consider multinomial logits for
nominal data, and ordered logit models for ordinal data, with a
brief mention of alternative-specific conditional logit models. We
will also consider sequential logit models.
(In line with the current syllabus we are skipping log-linear
models for contingency tables, and thus their relationship with
multinomial logit models.)

## 6.1 The Nature of Multinomial Data

We start by reading the data on contraceptive choice by age for currently
married women in El Salvador, 1985, found in Table 6.1 of the lecture notes.
The data are in "long" format with one row for each combination of predictor
and response, showing the age group, method choice, and number of cases.
In R we reshape the data so each method choice is a column, a layout that
works better with the functions we will use.

```{r}
library(haven)
library(tidyr)
cuselong <- read_dta("elsalvador1985.dta")
cuse <- pivot_wider(cuselong, names_from=cuse, values_from=cases)
names(cuse)[2:4] <- c("ster", "other", "none")
cuse
```

With only one predictor, this example affords limited opportunities for
interpreting coefficients, but will allow us to focus on the outcome and
the comparisons underlying each type of model.

<small>Updated fall 2022</small>
