
```{r include=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE)
```

## 8.3 Longitudinal Logits

We will work with a dataset on union membership used in the Stata manuals
and in my own paper on intra-class correlation for binary data. This is
a subsample of the National Longitudinal Survey of Youth (NLSY) and has
union membership information from 1970-88 for 4,434 women aged 14-26 in
1968. The data are available in the Stata website

```{r}
library(haven)
union = read_dta("https://www.stata-press.com/data/r14/union.dta")
```

### Logits

Here is a logit model

```{r}
logit <- glm(union ~ age + grade + not_smsa + south*year, 
  data = union, family = binomial)
summary(logit)
```

### Fixed-Effects

Let us try a fixed-effects model first.
My preferred way to fit this model in R is using the `clogit` function in 
the `survival` package, which requires specifying the group as `strata()`. 
Alternatives are the packages `gplm` and `glmmML`. I was able to verify that 
I get exactly the same results with `glmmboot()` in the `glmmML` package.

```{r}
library(survival)
fe <- clogit(union ~ age + grade + not_smsa + south*year + 
  strata(idcode) , data=union)
summary(fe)
library(dplyr)
group_by(union, idcode) |> summarize(v=var(union)) |>
  filter(v > 0) |> nrow()
```

It is not immediately obvious from the output that we lost 62% of our sample,
down from 4434 to 1690 women.
The 2744 drop outs are women who didn't have variation in union membership over
time. We will compare the estimates later.

### Random-Effects

Now we fit a random-effects model:

I tried to fit this model using `glmer()` in the `lme4` package, but it didn't 
converge, either with the default settings, which uses the Laplace approximation, 
nor when I specified 8-point adaptive quadrature. In all fairness this is a tough 
dataset, with very high intra-woman correlation.

I then tried the `glmmML()` function in the package of the same name, for which 
the default family is binomial, and it converged using the default Laplace 
approximation, as well as 8-point and 12-point Gauss-Hermite quadrature. 
(Adaptive Gauss Hermite is not available.)

```{r}
library(glmmML)
model = union ~ age + grade + not_smsa + south*year
re1  <- glmmML(model, cluster = idcode, data = union)
re8  <- glmmML(model, cluster = idcode, data = union, method = "ghq")
re12 <- glmmML(model, cluster = idcode, data = union, method = "ghq", n.points=12)
cbind(coef(re1), coef(re8), coef(re12))
re12
```		  

### Comparisons

Here's a table comparing the estimates for logit, random and fixed-effects models.

```{r}
cbind(coef(logit), coef(re12), c(NA,coef(fe)))
```

The main change is in the coefficient of `not_smsa`. You might think
this indicates something wrong with the logit and random-effects models,
but note that only women who have *moved* between standard metropolitan
statistical areas and other places contribute to the fixed-effects
estimate. It seems reasonable to believe that these women differ from
the rest.

The random-effect coefficients are larger in magnitude than the ordinary
logit coefficients. This is almost always the case. Omission of the
random effect biases the coefficients towards zero.

### Intra-class Correlation

The scale parameter of 2.395, available as `sigma`, can be translated into
an intraclass correlation as shown below:

```{r}
re12$sigma^2/(re12$sigma^2 + pi^2/3)
```

The intra-class correlation of 0.636 in the random-effects model indicates
a high correlation between a woman's propensity to be a union member
in different years, after controlling for education and residence.

My paper with Elo in the Stata journal shows how this coefficient can be
interpreted in terms of an odds ratio, and translated into measures of manifest
correlation such as Pearson's r and Yule's Q.

### Reference

Rodríguez, G. and Elo, I. (2003). "Intra-class correlation in random-effects
models for binary data". *The Stata Journal*,__3__(1):32-46.
<https://journals.sagepub.com/doi/pdf/10.1177/1536867X0300300102>

<small>Updated fall 2022</small>
