
```{r include=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE)
```

## 6.A Interpreting Multinomial Logit Coefficients {#mlogit .first}

Let us consider Example 16.1 in Wooldridge (2010), concerning school and
employment decisions for young men. The data contain information on
employment and schooling for young men over several years. We will work
with the data for 1987.

The outcome is status, coded 1=in school, 2=at home (meaning not in
school and not working), and 3=working. The predictors are education, a
quadratic on work experience, and an indicator for black.

We read the data from the Stata website, keep the year 1987, drop
missing values, label the outcome, and fit the model.
The argument `reltol` is used to increase the precision of the estimates

```{r}
library(haven)
library(dplyr)
library(nnet)
keane <- read_dta("https://www.stata.com/data/jwooldridge/eacsap/keane.dta")
keane <- filter(keane, year==87, !is.na(status)) |>
	mutate(status=factor(status, labels=c("school","home","work")))	
ml <- multinom(status ~ educ + exper + expersq + black, data=keane, reltol=1e-12)
summary(ml)
```

The results agree exactly with Table 16.1 in Wooldridge (2010, page 645).

### Relative Probabilities

Let us focus on the coefficient of black in the work equation, which is
0.311. Exponentiating we obtain

Thus, the relative probability of working rather than being in school is
37% higher for blacks than for non-blacks with the same education and
work experience. (Relative probabilities are also called relative odds.)

A common mistake is to interpret this coefficient as meaning that the
probability of working is higher for blacks. It is only the *relative*
probability of work over school that is higher. To obtain a fuller
picture we need to consider the second equation as well. The coefficient
of black in the home equation is 0.813. Exponentiating, we obtain

Thus, the relative probability of being at home rather than in school
for blacks is more than double the corresponding relative probability
for non blacks with the same education and work experience.

In short, black is associated with an increase in the relative
probability of work over school, but also a much large increase in the
relative probability of home over school. What happens with the actual
probability of working depends on how these two effects balance out.

### Marginal Effects (Continuous)

To determine the effect of black in the probability scale we need to
compute marginal effects, which can be done using continuous or discrete
calculations.

The continuous calculation is based on the derivative of the probability
of working with respect to a predictor. Let $\pi_{ij}=\Pr\{Y_i=j\}$
denote the probability that the i-th observation
follows on the j-th category, which is given by
$$
\pi_{ij} = \frac{e^{x_i'\beta_j}}{\sum_r e^{x_i'\beta_r}}
$$
where $\beta_j = 0$ when j is the baseline or reference outcome, in this
case school.

Taking derivatives w.r.t. the k-th predictor we obtain, after some
simplification
$$
 frac{\partial\pi\_{ij}}{\partial x_{ik}} =
\pi_{ij} ( \beta\_{jk} - \sum_r \pi_{ir} \beta_{rk} )
$$
noting again that the coefficient is zero for the baseline outcome.

To compute these we predict the probabilities and then apply the
formula.

```{r}
b <- coef(ml)
pr <- data.frame(predict(ml, type="p"))
part <- pr$home*b["home","black"] + pr$work*b["work","black"]
me <- data.frame( 
  school = pr$school * (                   - part),
  home   = pr$home   * (b["home", "black"] - part),
  work   = pr$work   * (b["work", "black"] - part))
summarize(me, school = mean(school), home = mean(home), work=mean(work))

```
We find that the average marginal effect of black on work is actually
negative: -0.0406. This means that the probability of working is on
average about four percentage points lower for blacks than for
non-blacks with the same education and experience.

### Marginal Effects (Discrete)

For the discrete calculation we compute predicted probabilities by
setting ethnicity to black and then to non-black and averaging:

```{r}
black    <- colMeans(predict(ml, mutate(keane, black=1), type="p"))
black
notblack <- colMeans(predict(ml, mutate(keane, black=0), type="p"))
notblack
data.frame(avg.marginal.effect = black["work"] - notblack["work"])
```

We find that the average probability of working is 0.7275 if black and
0.7710 if not black, a difference of -0.0435, so the probability of
working is on average just over four percentage points lower for blacks.

The take away conclusion here is that multinomial logit coefficients can
only be interpreted in terms of relative probabilities. To reach
conclusions about actual probabilities we need to calculate continuous
or discrete marginal effects.

### Reference

Wooldridge, J. M. (2010). *Econometric Analysis of Cross Section and Panel
Data*. Second Edition. Cambridge, Massachussets: The MIT Press.

<small>Updated fall 2022</small>
