
```{r include=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE)
```

## 7.A Piecewise Exponential Models

This is an illustration of piecewise exponential survival models using
individual-level data.
We use R, relying on the functions `survSplit()` to create pseudo-observations 
and `glm()` to fit the model using the Poisson equivalence. I also use the 
package `dplyr` for data manipulation. The call to `survSplit()` now uses a
formula interface. Hat tip to Jack Hoskins, who alerted me to this change 
starting with version 2.40-1.

### The Data

The dataset we will consider is analyzed in Wooldridge (2002) and
credited to Chung, Schmidt and Witte (1991). The data pertain to a
random sample of convicts released from prison between July 1, 1977 and
June 30, 1978. Of interest is the time until they return to prison. The
information was collected retrospectively by looking at records in April
1984, so the maximum length of observation is 81 months. The data are
available from the Stata website in Stata format.

```{r}
library(haven)
recid <- read_dta("https://www.stata.com/data/jwooldridge/eacsap/recid.dta")
nrow(recid)
```

The file has a censoring indicator, which we subtract from 1 to get a
failure indicator. We also create an id variable and list observation
number 9, which goes back to prison after 54 months.

```{r}
library(dplyr)
recid <- mutate(recid, fail = 1 - cens, id = row_number())
filter(recid, id == 9) |> select(id, durat, fail)
```

### Creating Pseudo-Observations

To create pseudo-observations for survival analysis we will use the 
`survSplit()` function in the `survival` package. We will split the data 
into single-year intervals of duration, from 0-12 to 48-60, with an open-ended 
category 60+. The function codes the interval variable using integer codes, 
and we turn that into a factor for convenience. We calculate exposure time 
for each episode as the difference between duration at the start and end.
We list these data for individual 9 to illustrate how the episodes are created.

```{r}
library(survival)
breaks <- seq(12, 60, by=12)
recidx <- survSplit(Surv(durat, fail) ~ ., data = recid, 
  cut = breaks, episode = "interval", start = "start")
recidx <- mutate(recidx, exposure = durat - start,
interval = factor(interval,  labels = 
  paste("(", c(0,breaks), ",", c(breaks,100), "]", sep=""))) |>
  rename(events = fail)
nrow(recidx)
filter(recidx, id==9) |> select(id,start,durat,interval, events, exposure)
```

The sample observation, which goes back to prison after 54 months, contributes
five episodes or pseudo-observations; one each for years one to four, with
12 months of exposure and no events, and another one for year five,  with
6 months of exposure and one event.

### A PWE Proportional Hazards Model

We are now ready to fit a proportional hazards model with a piecewise
exponential baseline where the hazard changes from year to year. We use
the same model as Wooldridge(2002), involving ten predictors, all fixed
covariates.
I specify the offset using the `exposure()` option. I could, of course,
take logs and then use the `offset()` option.]{.stata}

```{r}
fit <- glm(events~interval+workprg+priors+tserved+felon+alcohol+drugs+
  black+married+educ+age+offset(log(exposure)), 
  data=recidx, family=poisson)
summary(fit)
1 - exp(coef(fit)["felon"])
```

We see that the risk of recidivism is about the same in the first two years,
but then decreases substantially with duration since release. At any given
duration, felons have 25% lower risk of recidivism than non-felons with the
same observed characteristics.
Subjects imprisoned for alcohol or drug related offenses have much higher risk
of recidivism, everything else being equal.

### Survival Probabilities

We now illustrate the calculation of survival probabilities, starting
with the baseline hazard.
[There are different ways to do these calculations in Stata, but I
will proceed from first principles using Mata.]{.stata}.
We will retrieve the coefficients, add the constant and the age effects
to obtain the log hazard, exponentiate to obtain hazards, multiply by 12
and sum to obtain the cumulative baseline hazard, and then exponentiate
to obtain the baseline survival. This is not particularly meaningful as
it would apply with all covariates set to zero, including age, which is
measured in months.

```{r}
b <- coef(fit)
h <- exp(b[1] + c(0, b[2:6]))
H <- cumsum(12*h)
S <- exp(-H)
names(S)[1] <- names(H)[1] <- "interval(0,12])"
S
```

We will now estimate the probability of staying out of prison for five
years given average values of the predictors. First we calculate the mean of
each predictor; we have to be careful to include only one observation per
person, so we restrict the data to the first interval.

```{r}
xvars <- names(coef(fit))[7:16]
pset <- filter(recidx, interval == "(0,12]") |>  select(xvars) 
means <- colMeans(pset)
```

Now that we have the means, we multiply each by the corresponding
coefficient to obtain the linear predictor `xb`, exponentiate to
obtain a relative risk, multiply by the baseline hazard, and then
calculated the predicted survival.

```{r}
xb <- sum(coef(fit)[xvars] * means)
exp(-(H * exp(xb)))[5]
```

Thus, the probability of staying out of prison for five years for the average
person is 65.7%.

We now calculate this probability for felons and non-felons, keeping all other
variables at their means. All we need to do is subtract from `xb` the coefficient
of felon times the mean, which gives the linear predictor for a non-felon. We
then add the coefficient of felon to get the linear predictor for a felon. In
both cases the other variables stay at their means.

```{r}
xb0 <- xb - coef(fit)["felon"] * means["felon"]
xb1 <- xb0 + coef(fit)["felon"]
data.frame(non.felon = exp(-H[5]*exp(xb0)), felon= exp(-H[5]*exp(xb1)))
```

The predicted probability is 70.8% for felons and 63.2% for non-felons
when all other characteristics are set to the mean, a difference of 7.6
percentage points. This is a marginal effect at the means.

An alternative calculation sets every person to be a felon or non-felon,
leaving all other characteristics as they are, and then averages the
predicted probability of surviving five years without returning to
prison.

```{r}
xb0 <- as.matrix(mutate(pset, felon=0)) %*% coef(fit)[xvars]
xb1 <- as.matrix(mutate(pset, felon=1)) %*% coef(fit)[xvars]
data.frame(non.felon=mean(exp(-H[5] * exp(xb0))), 
               felon=mean(exp(-H[5] * exp(xb1))))
```

The average probability of staying out of prison for five years is 68.6%
for felons 61.2% for non-felons, a difference of 7.4 percentage points. This
can be interpreted as an average marginal effect.

### References

Wooldridge, Jeffrey M. (2010). *Econometric Analysis of Cross Section and Panel
Data*. 2nd Edition. Cambridge, Massachusetts: The MIT Press.

Chung, C-F, P. Schmidt and A.D. Witte (1991). "Survival Analysis: A Survey".
*Journal of Quantitative Criminology*,__7__:59-98.

<small>Updated fall 2022</small>
