
# 5 Generalized Linear Models

Generalized linear models are just as easy to fit in R as ordinary linear 
model. In fact, they require only an additional parameter to specify the 
variance and link functions.

## 5.1 Variance and Link Families

The basic tool for fitting generalized linear models is the `glm()` function, 
which has the folllowing general structure:

```{.r}
> glm(formula, family, data, weights, subset, ...)
```

where `...` stands for more esoteric options. The only parameter that we have 
not encountered before is `family`, which is a simple way of specifying a 
choice of variance and link functions. There are six choices of family:

Family	          Variance          Link
----------------- ----------------- -----------
gaussian	        gaussian          identity
binomial	        binomial	        logit, probit or cloglog
poisson           poisson	          log, identity or sqrt
gamma             gamma	            inverse, identity or log
inverse.gaussian	inverse.gaussian  $1/\mu^2$
quasi	            user-defined	    user-defined

As can be seen, each of the first five choices has an associated variance 
function (for binomial, the binomial variance $\mu(1 - \mu)$, and one or more 
choices of link functions (for binomial, the logit, probit or complementary 
log-log links).

As long as you want the default link, all you have to specify is the family 
name. If you want an alternative link, you must add a link argument. For 
example to do probits you use

```{.r}
> glm( formula, family = binomial(link = probit))
```

The last family on the list, `quasi`, is there to allow fitting user-defined 
models by maximum quasi-likelihood.

## 5.2 Logistic Regression

We will illustrate fitting logistic regression models using the contraceptive 
use data excerpted below (and shown in full further below):

```{.r}
  age education wantsMore notUsing using 
  <25       low       yes       53     6
  <25       low        no       10     4
  ... lines omitted ...
40-49      high        no       12    31
```
The data are available from the datasets section of the website for my 
generalized linear models course. Visit 
<https://data.princeton.edu/wws509/datasets> to read a short description and 
follow the link to `cuse.dat`.

Of course the data can be downloaded directly from within R:

```{r}
cuse = read.table("https://data.princeton.edu/wws509/datasets/cuse.dat", header 
= TRUE)
cuse
```

I specified the `header` parameter as `TRUE`, because otherwise it would not 
have been obvious that the first line in the file has the variable names. There 
are no row names specified, so the rows will be numbered from 1 to 16. I also 
printed the data to make sure we got it alright.

Let us try a simple additive model where contraceptive use depends on age, 
education and whether or not the woman wants more children:

```{r}
lrfit = glm( cbind(using, notUsing) ~ age + education + wantsMore, 
  data = cuse, family = binomial)
```

There are a few things to explain here. First, the function is `glm()` and I 
have assigned its value to an object called `lrfit` (for logistic regression 
fit). The first argument of the function is a model formula, which defines the 
response and linear predictor.

With binomial data the response can be either a vector or a matrix with two 
columns.

* If the response is a *vector*, it can be numeric with 0 for failure and 1 for 
success, or a factor with the first level representing "failure" and all others 
representing "success". In these cases R generates a vector of ones to 
represent the binomial denominators.

* Alternatively, the response can be a *matrix* where the first column is the 
number of "successes" and the second column is the number of "failures". In 
this case R adds the two columns together to produce the correct binomial 
denominator.

Because the latter approach is clearly the right one for us, I used the 
function `cbind()` to create a matrix by binding the column vectors containing 
the numbers using and not using contraception.

Following the special symbol `~` that separates the response from the 
predictors, we have a standard Wilkinson-Rogers model formula. In this case we 
are specifying main effects of `age`, `education` and `wantsMore`. Because all 
three predictors are categorical variables, they are treated automatically as 
factors, as you can see by inspecting the results:
  
``` {r}
lrfit
```

Recall that R sorts the levels of a factor in alphabetical order. Because `<25` 
comes before `25-29`, `30-39`, and `40-49`, it has been picked as the reference 
cell for `age`. Similarly, `high` is the reference cell for `education` because 
high comes alphabetically before `low`! Finally, R picked `no` as the base for 
`wantsMore`.

If you are unhappy about these choices you can (1) use `relevel()` to change 
the base category, or (2) define your own indicator variables. I will use the 
second approach,  defining indicators for women with high education and women 
who want no more children, both added to the `cuse` data frame:

```{r}
cuse$noMore = cuse$wantsMore == "no"
cuse$hiEduc = cuse$education == "high"
```

Now try the model with these predictors

```{r}
glm(cbind(using, notUsing) ~ age + hiEduc + noMore, 
  data = cuse, family = binomial)
```


The residual deviance of 29.92 on 10 d.f. is highly significant:

```{r}
pchisq(29.92, 10, lower.tail = FALSE)
```

To obtain a p-value I specified `lower.tail` as `FALSE`. This is more accurate 
than computing the default lower tail and subtracting from one.

So, we need a better model. One of my favorites for this dataset introduces an 
interaction between age and wanting no more children,
which is easily specified

```{r}
lrfit2 = glm( cbind(using, notUsing) ~ age * noMore + hiEduc , data = cuse, 
family = binomial)
lrfit2
```

Note how R built the interaction terms automatically, and even came up with 
sensible labels for them. The model's deviance of 12.63 on 7 d.f. is not 
significant at the conventional five per cent level, so we have no evidence 
against this model.

To obtain more detailed information about this fit try the `summary()` function:

```{r}
summary(lrfit2)
```

R follows the popular custom of flagging significant coefficients with one, two 
or three stars depending on their p-values. Try `plot(lrfit2)`. You get the 
same plots as in a linear model, but adapted to a generalized linear model; for 
example the residuals plotted are deviance residuals (the square root of the 
contribution of an observation to the deviance, with the same sign as the raw 
residual).

The functions that can be used to extract results from the fit include

- `residuals()` or `resid()`, for the deviance residuals
- `fitted()` or `fitted.values()`, for the fitted values (estimated 
probabilities)
- `predict()`, for the linear predictor (estimated logits)
- `coef()` or `coefficients()`, for the coefficients, and
- `deviance()`, for the deviance.

Some of these functions have optional arguments; for example, you can extract 
five different types of residuals, called "deviance", "pearson", "response" 
(defined as response - fitted value), "working" (the working dependent variable 
in the IRLS algorithm - linear predictor), and "partial" (a matrix of working 
residuals formed by omitting each term in the model). You specify the one you 
want using the `type` argument, for example `residuals(lrfit2, type = 
"pearson")`.

## 5.3 Model Updating

If you want to modify a model you may consider using the special function 
`update()`. For example to drop the `age:noMore` interaction in our model, one 
could use

 ```{r}
 lrfit1 = update(lrfit2, ~ . - age:noMore)
 ```
 
The first argument is the result of a fit, and the second an updating formula. 
The tilde `~` separates the response from the predictors, and the dot `.` 
refers to the right-hand side of the original formula, so here we simply remove 
`age:noMore`. Alternatively, one can give a new formula as the second argument.
 
The update function may also be used to fit the same model to different 
datasets, using the argument `data` to specify a new data frame. Another useful 
argument is `subset`, to fit the model to a different subsample. This function 
works with linear models as well as generalized linear models.

If you plan to fit a sequence of models you will find the anova function 
useful. Given a series of nested models, it will calculate the change in 
deviance between them. Try

```{r}
anova(lrfit1, lrfit2)
```

Adding the interaction has reduced the deviance by 17.288 at the expense of 3 
d.f.

If the argument to `anova()` is a single model, the function will show the 
change in deviance obtained by adding each of the terms in the order listed in 
the model formula, just as it did for linear models. Because this requires 
fitting as many models as there are terms in the formula, the function may take 
a while to complete its calculations.

The `anova()` function lets you specify an optional test. The usual choices 
will be "F" for linear models and "Chisq" for generalized linear models. Adding 
the parameter `test = "Chisq"` adds p-values next to the deviances. In our case

```{r}
anova(lrfit2, test = "Chisq")
```

We can see that all terms were highly significant when they were introduced 
into the model.

## 5.4 Model Selection

A very powerful tool in R is a function for stepwise regression that has three 
remarkable features:

* It works with generalized linear models, so it will do stepwise logistic 
regression, or stepwise Poisson regression,

* It understand hierarchical models, so it will only consider adding 
interactions after including the corresponding main effects in the models, and

* It understands terms involving more than one degree of freedom, so it it will 
keep together dummy variables representing the effects of a factor.

The basic idea of the procedure is to start from a given model (which could 
well be the null model) and take a series of steps, by either deleting a term 
already in the model, or adding a term from a list of candidates for inclusion, 
called the scope of the search and defined, of course, by a model formula.

Selection of terms for deletion or inclusion is based on Akaike's information 
criterion (AIC). R defines AIC as

AIC = –2 maximized log-likelihood + 2 number of parameters

The procedure stops when the AIC criterion cannot be improved.

In R all of this work is done by calling a couple of functions, `add1()` and 
`drop1()~, that consider adding or dropping one term from a model. These 
functions can be very useful in model selection, and both of them accept a 
`test` argument just like `anova()`.

Consider first `drop1()`. For our logistic regression model,

```{r}
drop1(lrfit2, test = "Chisq")
```

Obviously we can't drop any of these terms. Note that R considered dropping the 
main effect of education, and the age by want no more interaction, but did not 
examine the main effects of age or want no more, because one would not drop 
these main effects while retaining the interaction.

The sister function `add1()` requires a scope to define the additional terms to 
be considered. In our example we will consider all possible two-factor 
interactions:

```{r}
add1(lrfit2, ~ .^2, test = "Chisq")
```

We see that neither of the missing two-factor interactions is significant by 
itself at the conventional five percent level. (However, they happen to be 
jointly significant.) Note that the model with the age by education interaction 
has a lower AIC than our starting model.

The `step()` function will do an automatic search. Here we let it start from 
the additive model and search in a scope defined by all two-factor interactions.


```{r results = FALSE}
search = step(lrfit1, ~.^2)
```

The `step()` function produces detailed trace output that I have supressed. The 
returned object, however, includes an `anova` component that summarizes the 
search:

```{r}
search$anova
```
As you can see, the automated procedure introduced, one by one, all three 
remaining two-factor interactions, to yield a final AIC of 99.9. This is an 
example where AIC, by requiring a deviance improvement of only 2 per parameter, 
may have led to overfitting the data.

Some analysts prefer a higher penalty per parameter. In particular, using 
`log(n)` instead of 2 as a multiplier yields BIC, the Bayesian Information 
Criterion. In our example `log(1607) = 7.38`, so we would require a deviance 
reduction of 7.38 per additional parameter. The `step()` function accepts `k` 
as an argument, with default 2. You may verify that specifying `k = log(1607)` 
leads to a much simpler model; not only are no new interactions introduced, but 
the main effect of education is dropped (even though it is significant). 

In this example AIC would lead to a model that may be too complex, and BIC 
would lead to a model that may be too simple. In my opinion, the model with 
only one interaction is just right.
